Bayesian Rate Estimation

### 2012-11-09 many.tex
Edits complete through SIIIB.

The task of estimating a rate of foreground events when a mixture of
foreground and background events is present in the data is a common
one in physical and astrophysical applications.  This problem comes
up, among others, in gravitational-wave data analysis
\cite[e.g.,][]{s5-highmass,s6-lowmass} and in astronomical
observations of a field of objects of mixed provenance \cite[GC
  cluster paper]{}..  In this paper, we introduce a robust formalism
for estimating event rates from the data when the shape of foreground
and background distributions are known (but may have additional free
parameters), but the provenance of individual events as either
background or foreground is unknown.

@@

We assume that we are presented with a data set of $N$ events that
exceed a pre-specified threshold in ranking statistic, $x_{\rm min}$.
Each event may be due to either a signal of interest or an
uninteresting background.  Each event is associated with a ranking
statistic, $x$.  Our data set therefore consists of the ranking
statistics for the set of events:
\begin{equation}
  d = \mathset{ x_i | i = 1, \ldots, N } .
\end{equation}
The number of events $N$ is also part of the observed data, but we
separate out $N$ and the observed ranking statistics, $d$, for
convenience. We can choose how to label our events. Ultimately we will
label the events in order of ranking statistic, i.e., $x_1 < x_2 <
\cdots < x_N$, but some of the derivations that follow are simpler if
the events are ordered by time of arrival. We will use $d$ to denote
ranking statistic-ordered events, and $d_{\rm to}$ to denote
time-ordered events.

@@

Not surprisingly, this ranking-statistic ordered posterior can be
computed directly by assuming that the flags, $\mathset{f_i}$, are
un-observed data and treating the sets $\mathset{x_i | f_i = 1}$ and
$\mathset{x_i | f_i = 0}$ as samples from an inhomogeneous Poisson
process.  For an inhomogeneous Poisson process with rate function
$r(y)$ (cumulative rate $R(y)$), the likelihood of a set of samples
$\mathset{y_i}$ is given by
\begin{multline}
  p\left( \mathset{y_i} | r \right) \,{\rm d}^Ny_i= P\left(
  \textnormal{zero events below } y_1 \right) \times P\left(
  \textnormal{one event between } y_1 \textnormal{ and } y_1 + {\rm
    d}y_1 \right) \\ \times P\left( \textnormal{zero events between }
  y_1+{\rm d}y_1 \textnormal{ and } y_2 \right) \ldots \\ p\left(
  \mathset{y_i} | r \right) = \lim_{\delta y_i \to 0} \exp\left[ -
    R\left(y_1\right) \right] \left[ r\left( y_1 \right) +
    \order{\delta y_1} \right] \times \exp\left[ - \left[ R\left( y_2
      \right) - R\left( y_1 + \delta y_1\right) \right] \right] \times
  \ldots \\ = \left[\prod_i r\left( y_i \right)\right] \exp\left[ -
    R\left( \infty \right) \right].
\end{multline}
Applying this once to the foreground samples, once to the background
samples and taking the product, we obtain $p(d,\mathset{f_i}, N | R_f,
R_b, \theta)$ and thence $p(\mathset{f_i}, R_f, R_b, \theta | d,N) =
p(d,\mathset{f_i}, N | R_f, R_b, \theta)\ p(R_f, R_b, \theta)\, /\,
p(d,N)$. With the identification $p(d,N) = N!\,p(d_{\rm to},N)$, as
justified above, we reproduce Eq.~\eqref{eq:posterior}.

@@

For the usual $k=1$ case \citep{Biswas2009}, the marginalised
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising
over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}\,
  p(\theta)}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)
  +
  \sqrt{\left(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)\right)^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta))}.
\end{equation}
where $\tilde{b}(x_N,\theta)
=\hat{b}(x_N,\theta)/(1-\hat{B}(x_N,\theta))$. If
$\tilde{b}(x_N,\theta) \ll 1$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be understood as the
statement that the rate of foreground events with ranking statistic
greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of order $1$,
as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term will be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) >> \hat{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavoured.

@@

\subsubsection{Foreground dominated statistic}
If we set the threshold for including an event, $x_{\rm min}$,
sufficiently high, we can ensure that $\hat{f}(x_i,\theta) \gg
\hat{b}(x_i,\theta)$ for all ranking statistics $x_i$ in the data
set. The posterior can then be approximated by
\begin{equation}
p_{\rm FD}\left( R_f, R_b, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^N \exp\left[-\left( R_f +
  R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
Normalisation over $R_b$ gives a constant factor and the posterior on
the foreground rate becomes
\begin{equation}
p_{\rm FD}\left( R_f, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^{N-\frac{1}{2}}
\exp\left[-R_f \right] p(\theta).
\end{equation}
Ignoring the dependence on $\theta$, this is peaked at a rate $R_f =
N-1/2$, so we have the expected result that, in the foreground
dominated regime, the rate is approximately equal to the number of
events observed. We note that the rate $R_f$ is defined as the rate of
events occurring above the specified threshold and therefore if the
foreground dominated statistic were used to estimate rates $R_{f,1}$
and $R_{f,2}$ for thresholds $x_{\rm min}=x_1$, $x_{\rm min}=x_2$
respectively, these rates should be compared by equating $R_{f,1}
(1-\hat{F}(x_2,\theta))$ and $R_{f,2} (1-\hat{F}(x_1,\theta))$.

@@

For a data stream of pure noise, $d(f) = n(f)$, the SNR of a each
template follows a $N(0,1)$ distribution.  The background ranking
statistic (i.e.\ the maximum SNR over the template bank) then has a
cumulative distribution without thresholding of
\begin{equation}
  \hat{B}(x) = \left( \frac{1 + \erf\left( \frac{x}{\sqrt{2}}
    \right)}{2} \right)^N
\end{equation}
where ${\rm erf}(x)$ is the error function as before. Imposing the
threshold, $x > \xmin$, the cumulative distribution of the background
becomes
\begin{equation}
  \label{eq:analytic-background-rate}
  \hat{B}(x) = \frac{\left( 1 + \erf\left( \frac{x}{{\sqrt{2}}}
    \right) \right)^N - \left( 1 + \erf\left( \frac{\xmin}{{\sqrt{2}}}
    \right) \right)^N}{2^N - \left( 1 + \erf\left(
    \frac{\xmin}{{\sqrt{2}}} \right) \right)^N }
\end{equation}
for $x>\xmin$, $0$ otherwise.

@@

To demonstrate the effectiveness of our formalism, we applied it to a
synthetic data set with foreground and background distributions drawn
from Eqs.~\eqref{eq:analytic-background-rate} and
\eqref{eq:analytic-foreground-rate} with $R_f^\mathrm{true} = 10.4$
and $R_b^\mathrm{true} = 95.1$, using 1000 templates.  \ilya{[Specify
    $\xmin$?]} The synthetic data consisted of 13 foreground events
and 85 background events; the cumulative distribution for the ranking
statistic of the synthetic data appears in Figure
\ref{fig:analytic-data-cumulative}.  We used a Markov Chain Monte
Carlo simulation to draw samples of state flags and rates from the
joint posterior (Eq.~\eqref{eq:posterior}).

@@

We can compare these results to results obtained using the two
approximations described earlier, the loudest-event statistic and the
foreground-dominated statistic. The marginalised distribution for the
foreground rate using these alternatives are shown in
Figure~\ref{fig:altmeth}. In this case, the loudest event had
$x_N=9.5$\jon{(correct this)}. The loudest-event statistic depends on
a specification of the maximum, $R_{\rm max}$, for the background
rate. We show results for $R_{\rm max}=\infty$, i.e., the improper
prior, and $R_{\rm max}=10000$. The results for other reasonable
choices of $R_{\rm max} = 100, 1000, 100000$ etc. gave exactly the
same posterior, since $\hat{b}(x_N) R_{\rm max} \ll \hat{f}(x_N)$ for
all these choices and we are therefore in the regime where the
posterior is insensitive to $R_{\rm max}$. To apply the
foreground-dominated statistic we must specify a threshold above which
we assume all events are foreground. It is reasonable to do this based
on a specification for the relative probability of an event being
fore/background, $\hat{f}(x)/\hat{b}(x)=p_{\rm thresh}$. Setting
$p_{\rm thresh}=0.99$ gives $x_{\rm min}=4.07$ and there are $N=13$
events exceeding that threshold. Setting $p_{\rm thresh}=0.5$ gives
$x_{\rm min}=3.82$ and there are $N=15$ events exceeding that
threshold \jon{(correct these numbers using Will's data)}. We show
results for both of these choices of $x_{\rm min}$ in
Figure~\ref{fig:altmeth}. We also show similar results for two extreme
cases --- setting $x_{\rm min}=3.5$ so that we assume all $98$
observed events are foreground; and setting $x_{\rm min}=9$ so that
only the loudest event is classified as foreground.

The loudest event statistic with the improper prior gives, as
expected, a poor approximation to the foreground rate. The peak is
more accurately located when a prior maximum rate is defined, but the
distribution is much wider than using the full analysis described
here. This is to be expected as much of the information is being
thrown away. The foreground-dominated statistic gives a reasonable
approximation to the true foreground rate, and a distribution that is
not much wider than the full analysis, for the non-extreme choices of
$x_{\rm min}$. For $x_{\rm min}=9$, the foreground dominated statistic
has equivalent performance to the loudest event statistic, as we might
expect, while when $x_{\rm min}=3.5$, it performs poorly since we are
approximating the foreground rate by the total foreground plus
background rate. This indicates that, provided the threshold is chosen
appropriately, the foreground dominated statistic can perform quite
well at estimating the rate. The fact that it reproduces the posterior
from the full analysis so well is indicative of the fact that most of
the information about the foreground comes from the loudest
events. Indeed, for $x_{\rm min} \approx 4$, the set of events
consists mostly of foreground sources and that set contains a decent
number of events ($13$), so it is not surprising that this statistic
can estimate the foreground rate well. The choice of $x_{\rm min}$ was
motivated by knowledge of $\hat{b}(x,\theta)$ and $\hat{f}(x,\theta)$,
which we have in this case, but this might not be possible if the
parameters $\theta$ were simultaneously being estimated. The full
analysis naturally incorporates inference about the rate parameters
$\theta$ and the background rate $R_b$ along with the foreground rate
and incorporates maximum information from the data set and should
therefore lead to narrower posteriors in general.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters (error bars are one standard deviation).  True
    foreground events are in dark grey, background events in light
    grey.  Even though our method cannot identify the status of most
    events with confidence, it can still correctly estimate the rates
    (Figure \ref{fig:analytic-rate-recovery}).}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{AltMethodComp}
  \caption{\label{fig:altmeth}Posteriors on foreground rate obtained
    using alternative methods --- the loudest event statistic and the
    foreground dominated statistic. Results are shown for two
    different choices of the maximum a priori background rate, $R_{\rm
      max}$, in the case of the loudest event (``LE'') statistic, and
    for several different choices of the threshold, $x_{\rm min}$, in
    the case of the foreground dominated (``FD'') statistic. \ilya{I
      can see the figure on the screen, but can't print it...}}
\end{figure}

@@

Results from such an analysis appear in Figures \ref{fig:rates-nt} and
\ref{fig:ntemplates}.  We use the same parameters and data set as in
\S \ref{sec:analytic-GW-example}, with $R_f = 10.4$, $R_b = 95.1$, and
$N_\mathrm{eff} = 1000$, but now allow $N_\mathrm{eff}$ to be a
parameter of the background distribution, with a flat prior.  Both the
rates and the number of effective templates are recovered without
significant loss of accuracy relative to the fixed $N_\mathrm{eff}$
situation in \S \ref{sec:analytic-GW-example}.

If we consider the two alternative methods, the loudest event and
foreground dominated statistics, we will recover the same foreground
distributions as in the case with $N_{\rm eff}$ fixed, which are shown
in Figure~\ref{fig:altmeth}. This is because the parameter $N_{\rm
  eff}$ affects only the background distribution, to which the
foreground-dominated statistic is insensitive, and in the loudest
event case, after marginalisation over $N_{\rm eff}$ we find
$\int_0^{N_{\rm max}} \hat{b}(x_N,N_{\rm eff}) {\rm d}N_{\rm eff} \ll
3 N_{\rm max}/R_{\rm max} \hat{f}(x_N, N_{\rm eff})$ and so we are
still in the foreground-dominated regime in which the loudest event
tells us nothing about the background. Similarly, neither of these
alternative methods can inform us about the value of $N_{\rm eff}$, a
property of the background. \jon{Technically, if we allowed really
  large values of $N_{\rm eff}$ we would get into a regime where the
  loudest event could be background and we'd get something here. But,
  we would need $N_{\rm max} \sim 10^{20}$ and so I don't think we
  should consider these alternatives further in this case.} \ilya{On
  the other hand, this ties in with the comment at the end of the
  previous subsection about the difficulty of choosing $x_{min}$ for
  the foreground-dominated statistic when there are other parameters
  $\theta$ to be estimated simultaneously ($N_{eff}$ in this case), so
  perhaps it's worthwhile to add some comment about this?}

>>>>

The task of estimating a rate of foreground events when a mixture of
foreground and background events is present in the data is a common
one in physical and astrophysical applications.  This problem comes
up, among others, in gravitational-wave data analysis
\cite[e.g.,][]{s5-highmass,s6-lowmass} and in astronomical
observations of a field of objects of mixed provenance
\cite{DeGennaro2009}.  In this paper, we introduce a robust formalism
for estimating event rates from the data when the shape of foreground
and background distributions are known (but may have additional free
parameters), but the provenance of individual events as either
background or foreground is unknown.

@@

We first consider one-dimensional data, but will generalize to the
multidimensional case below.  We assume that we are presented with a
data set of $N$ events that exceed a pre-specified threshold in
ranking statistic, $x_{\rm min}$.  Each event may be due to either a
signal of interest or an uninteresting background.  Each event is
associated with a ranking statistic, $x$.  Our data set therefore
consists of the ranking statistics for the set of events:
\begin{equation}
  d = \mathset{ x_i | i = 1, \ldots, N } .
\end{equation}
The number of events $N$ is also part of the observed data, but we
separate out $N$ and the observed ranking statistics, $d$, for
convenience. We can choose how to label our events. Ultimately we will
label the events in order of ranking statistic, i.e., $x_1 < x_2 <
\cdots < x_N$, but some of the derivations that follow are simpler if
the events are ordered by time of arrival (i.e.\ randomly with respect
to the $x_i$). We will use $d$ to denote ranking statistic-ordered
events, and $d_{\rm to}$ to denote time-ordered events.

@@

The ranking-statistic ordered posterior can be computed directly by
assuming that the flags, $\mathset{f_i}$, are un-observed data and
treating the sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i =
  0}$ as samples from an inhomogeneous Poisson process.  For an
inhomogeneous Poisson process with rate function $r(y)$ (cumulative
rate $R(y)$), the likelihood of a set of samples $\mathset{y_i}$ is
given by
\begin{multline}
  p\left( \mathset{y_i} | r \right) \,{\rm d}^Ny_i= P\left(
  \textnormal{zero events below } y_1 \right) \times P\left(
  \textnormal{one event between } y_1 \textnormal{ and } y_1 + {\rm
    d}y_1 \right) \\ \times P\left( \textnormal{zero events between }
  y_1+{\rm d}y_1 \textnormal{ and } y_2 \right) \ldots \\ p\left(
  \mathset{y_i} | r \right) = \lim_{\delta y_i \to 0} \exp\left[ -
    R\left(y_1\right) \right] \left[ r\left( y_1 \right) +
    \order{\delta y_1} \right] \times \exp\left[ - \left[ R\left( y_2
      \right) - R\left( y_1 + \delta y_1\right) \right] \right] \times
  \ldots \\ = \left[\prod_i r\left( y_i \right)\right] \exp\left[ -
    R\left( \infty \right) \right].
\end{multline}
Applying this once to the foreground samples, once to the background
samples and taking the product, we obtain $p(d,\mathset{f_i}, N | R_f,
R_b, \theta)$ and thus $p(\mathset{f_i}, R_f, R_b, \theta | d,N) =
p(d,\mathset{f_i}, N | R_f, R_b, \theta)\ p(R_f, R_b, \theta)\, /\,
p(d,N)$. With the identification $p(d,N) = N!\,p(d_{\rm to},N)$, as
justified above, we reproduce Eq.~\eqref{eq:posterior}.

@@

For the usual $k=1$ case \citep{Biswas2009}, the marginalised
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising
over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}\,
  p(\theta)}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)
  +
  \sqrt{\left(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)\right)^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta))}.
\end{equation}
where $\tilde{b}(x_N,\theta) = \hat{b}(x_N,\theta) /
(1-\hat{B}(x_N,\theta))$\footnote{The quantity $\Lambda$ from
  \citet{Biswas2009} is related to $\hat{b}$ and $\tilde{f}$ by
  \begin{equation}
    \Lambda(x, \theta) = \frac{\tilde{f}(x,\theta)}{R_b \hat{b}(x,\theta)}.
\end{equation}}.  If $\tilde{b}(x_N,\theta) \ll 1$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be understood as the
statement that the rate of foreground events with ranking statistic
greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of order $1$,
as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term will be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) >> \hat{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavoured.

@@

\subsubsection{Foreground dominated statistic}
If we set the threshold for including an event, $x_{\rm min}$,
sufficiently high, we can ensure that $\hat{f}(x_i,\theta) \gg
\hat{b}(x_i,\theta)$ for all ranking statistics $x_i$ in the data
set. The posterior can then be approximated by
\begin{equation}
p_{\rm FD}\left( R_f, R_b, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^N \exp\left[-\left( R_f +
  R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
Normalisation over $R_b$ gives a constant factor and the posterior on
the foreground rate becomes
\begin{equation}
p_{\rm FD}\left( R_f, \theta | d \right) \propto \prod_{i} \left[
  \hat{f}\left(x_i, \theta\right)\right] R_f^{N-\frac{1}{2}}
\exp\left[-R_f \right] p(\theta).
\end{equation}
Ignoring the dependence on $\theta$, this is peaked at a rate $R_f =
N-1/2$, so we have the expected result that, in the foreground
dominated regime, the rate is approximately equal to the number of
events observed (the $1/2$ comes from our use of the Jeffreys prior on
the rate).  We note that the rate $R_f$ is defined as the rate of
events occurring above the specified threshold and therefore if the
foreground dominated statistic were used to estimate rates $R_{f,1}$
and $R_{f,2}$ for thresholds $x_{\rm min}=x_1$, $x_{\rm min}=x_2$
respectively, these rates should be compared by equating $R_{f,1}
(1-\hat{F}(x_2,\theta))$ and $R_{f,2} (1-\hat{F}(x_1,\theta))$.

@@

For a data stream of pure noise, $d(f) = n(f)$, the SNRs of the
templates are independent $N(0,1)$ random variables.  The background
ranking statistic (i.e.\ the maximum SNR over the template bank) then
has a cumulative distribution without thresholding of
\begin{equation}
  \hat{B}(x) = \left( \frac{1 + \erf\left( \frac{x}{\sqrt{2}}
    \right)}{2} \right)^N
\end{equation}
where ${\rm erf}(x)$ is the error function as before. Imposing the
threshold, $x > \xmin$, the cumulative distribution of the background
becomes
\begin{equation}
  \label{eq:analytic-background-rate}
  \hat{B}(x) = \frac{\left( 1 + \erf\left( \frac{x}{{\sqrt{2}}}
    \right) \right)^N - \left( 1 + \erf\left( \frac{\xmin}{{\sqrt{2}}}
    \right) \right)^N}{2^N - \left( 1 + \erf\left(
    \frac{\xmin}{{\sqrt{2}}} \right) \right)^N }
\end{equation}
for $x>\xmin$, $0$ otherwise.

@@

To demonstrate the effectiveness of our formalism, we applied it to a
synthetic data set with foreground and background distributions drawn
from Eqs.~\eqref{eq:analytic-background-rate} and
\eqref{eq:analytic-foreground-rate} using $x_\mathrm{min} = 3.5$, with
$R_f^\mathrm{true} = 10.4$ and $R_b^\mathrm{true} = 95.1$ and 1000
templates.  The synthetic data consisted of 13 foreground events and
85 background events; the cumulative distribution for the ranking
statistic of the synthetic data appears in Figure
\ref{fig:analytic-data-cumulative}.  We used a Markov Chain Monte
Carlo simulation to draw samples of state flags and rates from the
joint posterior (Eq.~\eqref{eq:posterior}).

@@

We can compare these results to results obtained using the two
approximations described earlier, the loudest-event statistic and the
foreground-dominated statistic. The marginalised distribution for the
foreground rate using these alternatives are shown in
Figure~\ref{fig:altmeth}. In this case, the loudest event had
$x_N\simeq 9.47$. The loudest-event statistic depends on a
specification of the maximum, $R_{\rm max}$, for the background
rate. We show results for $R_{\rm max}=\infty$, i.e., the improper
prior, and $R_{\rm max}=10000$. The results for other reasonable
choices of $R_{\rm max} = 100, 1000, 100000$ etc. gave exactly the
same posterior, since $\hat{b}(x_N) R_{\rm max} \ll \hat{f}(x_N)$ for
all these choices and we are therefore in the regime where the
posterior is insensitive to $R_{\rm max}$. To apply the
foreground-dominated statistic we must specify a threshold above which
we assume all events are foreground. It is reasonable to do this based
on a specification for the relative probability of an event being
fore/background, $\hat{f}(x)/\hat{b}(x)=p_{\rm thresh}$. Setting
$p_{\rm thresh}=0.99$ gives $x_{\rm min}=4.07$ and there are $N=18$
(11 foreground and 7 background) events exceeding that
threshold. Setting $p_{\rm thresh}=0.5$ gives $x_{\rm min}=3.82$ and
there are $N=30$ (11 foreground and 19 background) events exceeding
that threshold.  Each of these thresholds gives a biased estimate of
the rate because there are background events still above threshold.
The ``omniscient'' threshold of $x_\mathrm{min} = 4.38$ produces $N=7$
(7 foreground and 0 background) events in this data set, and therefore
an unbiased estimate, but of course this threshold can only be
determined because we can examine the synthetic foreground and
background data samples.  We show results for the first two choices of
$x_{\rm min}$ in Figure~\ref{fig:altmeth}; the omniscient choice
produces essentially the same posterior as our full analysis.

The loudest event statistic with the improper prior gives, as
expected, a poor approximation to the foreground rate. The peak is
more accurately located when a prior maximum rate is defined, but the
distribution is much wider than using the full analysis described here
in any case. This is to be expected as much of the information is
being thrown away. The foreground-dominated statistic gives a
reasonable approximation to the true foreground rate, and a
distribution that is essentially equal to the full analysis, for the
``omniscent'' choice of threshold value that excludes all background
data. For lower thresholds, even for a threshold where
$p_\mathrm{thresh} = 0.99$, it performs poorly since we are
approximating the foreground rate by the total foreground plus
background rate.  This indicates that, provided the threshold is
chosen appropriately, the foreground dominated statistic can perform
quite well at estimating the rate---but choosing this threshold
correctly is difficult. The fact that it reproduces the posterior from
the full analysis so well is indicative of the fact that most of the
information about the foreground comes from the loudest events. The
full analysis naturally incorporates inference about the background
rate $R_b$ along with the foreground rate and incorporates maximum
information from the data set and should therefore lead to narrower
posteriors in general.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters.  True foreground events are in dark grey, background
    events in light grey.  Even though our method cannot identify the
    status of most events with confidence, it can still correctly
    estimate the rates (Figure \ref{fig:analytic-rate-recovery}).}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{AltMethodComp}
  \caption{\label{fig:altmeth}Posteriors on foreground rate obtained
    using the method described in this paper, the loudest event
    statistic and the foreground dominated analysis for the data set
    from \S \ref{sec:analytic-GW-example}.  For the loudest event
    statistic, we present the posterior with and without an upper
    limit on the background rate, $R_b$; in both cases the rate
    posterior is significantly wider than the one obtained with the
    method described in this paper.  For the foreground dominated
    statistic, the limits $x_\mathrm{min} = 3.82$ and $x_\mathrm{min}
    = 4.07$ give likelihood ratios of $\hat{f}/\hat{b} = 0.5$ and
    $0.99$.  For this data set, the thresholds in fact include 19 and
    7 background events, respectively, so the corresponding rate
    estimates are significantly biased.  An ``omniscient'' threshold
    of $x_\mathrm{min} = 4.38$ would produce exactly 7 foreground and
    zero background events, and the resulting posterior is essentially
    indistinguishable from the curve for the full analysis.}
\end{figure}

@@

Results from such an analysis appear in Figures \ref{fig:rates-nt} and
\ref{fig:ntemplates}.  We use the same parameters and data set as in
\S \ref{sec:analytic-GW-example}, with $x_\mathrm{min} = 3.5$, $R_f =
10.4$, $R_b = 95.1$, and $N_\mathrm{eff} = 1000$, but now allow
$N_\mathrm{eff}$ to be a parameter of the background distribution,
with a flat prior.  Both the rates and the number of effective
templates are recovered without significant loss of accuracy relative
to the fixed $N_\mathrm{eff}$ situation in \S
\ref{sec:analytic-GW-example}.

If we consider the two alternative methods, the loudest event and
foreground dominated statistics, and apply the same
foreground-dominated thresholds as before, we will recover the same
foreground distributions as are shown in Figure~\ref{fig:altmeth}.
This is because the parameter $N_{\rm eff}$ affects only the
background distribution, to which the foreground-dominated statistic
is insensitive, and in the loudest event case, after marginalisation
over $N_{\rm eff}$ we find $\int_0^{N_{\rm max}} \hat{b}(x_N,N_{\rm
  eff}) {\rm d}N_{\rm eff} \ll 3 N_{\rm max}/R_{\rm max} \hat{f}(x_N,
N_{\rm eff})$ and so we are still in the foreground-dominated regime
in which the loudest event tells us nothing about the background.
Neither of these alternative methods can inform us about the value of
$N_{\rm eff}$, a property of the background.  Moreover, the choice of
threshold value for the foreground-dominated statistic becomes
significantly more complicated in this case, since $p_\mathrm{thresh}$
now depends on $N_\mathrm{eff}$.
